HIPPOCAMPAL SEGMENTATION WITH A VERY SMALL ATLAS LIBRARY: THE MAGeT BRAIN ALGORITHM

### 2012-11-01 paper.Rnw
Incorporated Mallar's changes, and fixed dangling references.

\title{HIPPOCAMPAL SEGMENTATION WITH A VERY SMALL ATLAS LIBRARY: THE MAGET
BRAIN ALGORITHM} 
\name{  Jon Pipitone$^{1}$ Jason P. Lerch$^{4,5}$ Miriam Friedel$^{4,5}$
        Aristotle Voineskos$^{1,3}$ Mallar Chakravarty$^{1,2,3}$ }

@@

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating region labels from manual
delineations to unlabeled images. Unfortunately, these approaches often rely on
a large number of such manual-segmentated atlases which take time and
significant expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
whilst still achieving similar accuracy to  other multi-atlas approaches.  We
perform repeated random subsampling validation on the IBSR dataset to compare
our approach to multi-atlas segmentation using the full IBSR dataset, and to
single-atlas (model-based) segmentation. Our results show that with only 8
input atlas, MAGeT brain can achieve to within 2.0\% segmentation accuracy of
the multi-atlas approach using 17 input atlases (mean $\kappa = 0.775$ vs.
$\kappa = 0.791$).

\end{abstract}

@@

Currently, the gold standard for neuroanatomical segmentation is manual
labelling by an expert human rater.  This is problematic for segmentation of
the hippocampus for several reasons.  First, manual segmentation takes a
significant investment of time and expertise \cite{Hammers2003} which may not
be readily available to researchers or clinicians.  Second, the amount of data
produced in neuroimaging experiments increasingly exceeds the capacity for
identification of specific neuroanatomical structures by an expert manual
rater.  Third, the true delination of hippocampal anatomy in MR images is
disputed\cite{Geuze2004}, as evidenced by efforts to create an unified
segmentation protocol\cite{Jack2011}.  Compounding each of these problems
is the significant neuroanatomical variability in the hippocampus
throughout the course of aging, development, and neuropsychiatric
disorders\cite{someone}.

@@

In previous work from our group \cite{Chakravarty2011}, we applied MAGeT brain
to the segmentation the striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. In the present work we extend our approach to the
hippocampus, a much more difficult structure to segment, and perform a thorough
validation over a range of atlas and template library sizes. Due to the small
number of atlases required, our method could easily accommodate different
hippocampal definitions. Our aim is not to improve on segmentation accuracy
beyond existing methods, but instead to provide a method that trades off manual
segmentation expertise for computational processing time whilst providing
sufficient accuracy for clinical and research applications.

@@

\subsection{Image Processing and Registration Method}
The N3 algorithm \cite{Sled1998} is first used to minimize the intensity
nonuniformity in each image.  Image registration is carried out in two phases.
In the first, a 12-parameter linear transformation (3 translations, rotations,
scales, shears) is estimated between images using an algorithm that maximizes
the correlation between blurred MR intensities and gradient magnitude over the
whole brain \cite{Collins}.  In the second phase, nonlinear registration is
completed using the ANIMAL algorithm \cite{Collins1995}: an iterative procedure
that estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These
larger deformations are then input to subsequent steps where the fit is refined
by estimating smaller deformations on data blurred with a Gaussian kernel with
a smaller FWHM. The final transformation is a set of local translations defined
on a bed of equally spaced nodes that were estimated through the optimization
of the correlation coefficient. For the purposes of this work we used the
regularization parameters optimized in Robbins et al.\cite{Robbins2004}. It
should be noted that the MAGeT brain algorithm is not dependant on this, or
any, particular choice of registration method\cite{Chakravarty2011}.

>>>>

\title{HIPPOCAMPAL SEGMENTATION WITH A VERY SMALL ATLAS LIBRARY: THE MAGeT
BRAIN ALGORITHM} 
\name{  Jon Pipitone$^{1}$ Jason P. Lerch$^{4,5}$ Miriam Friedel$^{4}$
        Aristotle N. Voineskos$^{1,3}$ M. Mallar Chakravarty$^{1,2,3}$ }

@@

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. Unfortunately, these
approaches often rely on a large number of such manually segmentated atlases
that take significant time and expertise to produce. We present an algorithm
for the automatic segmentation of the hippocampus that minimizes the number of
atlases needed whilst still achieving similar accuracy to multi-atlas
approaches.  We perform repeated random subsampling validation on the IBSR
dataset to compare our approach to multi-atlas segmentation using the full IBSR
dataset, and to single-atlas (model-based) segmentation. Our results show that
with only 8 input atlas, MAGeT brain can achieve to within 2.0\% segmentation
accuracy of the multi-atlas approach using 17 input atlases (mean $\kappa =
0.775$ vs.  $\kappa = 0.791$).  These results compare favourably to others who
have used the IBSR data for validation. 
\end{abstract}

@@

Currently, the gold standard for neuroanatomical segmentation is manual
labelling by an expert human rater.  This is problematic for segmentation of
the hippocampus for several reasons.  First, manual segmentation takes a
significant investment of time and expertise \cite{Hammers2003} which may not
be readily available to researchers or clinicians.  Second, the amount of data
produced in neuroimaging experiments increasingly exceeds the capacity for
identification of specific neuroanatomical structures by an expert manual
rater.  Third, the true delineation of hippocampal anatomy in MR images is
disputed\cite{Geuze2004}, as evidenced by efforts to create an unified
segmentation protocol\cite{Jack2011}.  Compounding each of these problems is
the significant neuroanatomical variability in the hippocampus throughout the
course of aging, development, and neuropsychiatric disorders\cite{Mouiha2011}.

@@

In previous work from our group \cite{Chakravarty2012}, we applied MAGeT brain
to the segmentation the striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. In the present work we extend our approach to the
hippocampus, a much more difficult structure to segment, and perform a thorough
validation over a range of atlas and template library sizes. Due to the small
number of atlases required, our method could easily accommodate different
hippocampal definitions. Our aim is not to improve on segmentation accuracy
beyond existing methods, but instead to provide a method that trades off manual
segmentation expertise for computational processing time whilst providing
sufficient accuracy for clinical and research applications.

@@

\subsection{Image Processing and Registration Method}
The N3 algorithm \cite{Sled1998} is first used to minimize the intensity
nonuniformity in each image.  Image registration is carried out in two phases.
In the first, a 12-parameter linear transformation (3 translations, rotations,
scales, shears) is estimated between images using an algorithm that maximizes
the correlation between blurred MR intensities and gradient magnitude over the
whole brain \cite{Collins}.  In the second phase, nonlinear registration is
completed using the ANIMAL algorithm \cite{Collins1995}: an iterative procedure
that estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These
larger deformations are then input to subsequent steps where the fit is refined
by estimating smaller deformations on data blurred with a Gaussian kernel with
a smaller FWHM. The final transformation is a set of local translations defined
on a bed of equally spaced nodes that were estimated through the optimization
of the correlation coefficient. For the purposes of this work we used the
regularization parameters optimized in Robbins et al.\cite{Robbins2004}. It
should be noted that the MAGeT brain algorithm is not dependant on this, or
any, particular choice of registration method\cite{Chakravarty2012}.
